Box

### 2021-07-22 MS/box.Rmd
Box updated with Losia's comments

Here, we show how to obtain sampling variance for a slope (or a 'rate of change') when you have measurements of a physiological trait at two points along an environmental gradient (e.g. temperature, salinity, acidity). We then derive sampling variance for the difference between two slopes, and demonstrate how SMD (standardized mean difference) can be corrected to account for differences in both units (e.g. cm or mg) and points on an environmental gradient (e.g. temperature of 24 and 30 \&degC). Finally, we introduce a useful approximation technique known as 'the delta method'. The sampling variances associated with these 'new' effect sizes below will allow comparative physiologists to take advantage of powerful meta-analytic models.

@@

Let's start with a real example that comparative physiologists can easily relate with, the slope of responses to temperature acclimation or,or as we call it, the acclimation response ratio (ARR). ARR can be defined as a slope for acclimated physiological responses at two different temperature points, defined as:

@@

Therefore, when $M_{1}$ and $M_{2}$ are independent, we can obtain the sampling variance for ARR as:

@@

where $SD_{1}$ and $SD_{2}$ and $N_{1}$ and $N_{2}$ are standard deviations and sampling sizes at temperature $T_{1}$ and $T_{2}$. Readers may find it difficult to see how we obtained this equation. Let us explain further. The (sampling) standard error for $X_{1}$ is $SE_{1} = SD_{1} / \sqrt{N_{1}}$. Given this, the sampling variance for $X_{1}$ is $SD^2_{1} / N_{1}$, and the sampling variance for $X_{2}$ is $SD^2_{2} / N_{2}$. You can see those elements of the sampling variance of ARR in the formula above. The term $(1/(T_{2} - T_{1}))^2$ comes from recognizing $1/(T_{2} - T_{1})$ as a constant.

In ecological and evolutionary studies, it is not uncommon that we measure the same group of organisms at two temperatures. If we do, however, then we need to add the covariance between $M_{1}$ and $M_{2}$ ($\text{Cov}(M_{1},M_{2})$) in the equation. Note that, as above, the covariance equals the correlation multiplied by the square-root of two (sampling) variances. Therefore, the sampling variance of ARR can be now written as:

@@

Now let us assume that we want to know the difference between two different ARR values: female ARR, $\text{ARR}_{f}$ and male ARR, $\text{ARR}_{m}$. Such a difference (ARRD) can be written as:

@@

Using the properties of variance and the equations from above, the sampling variance for ARRD can be derived, when measurements at two temperature points are independent, as:

@@

Here we have derived sampling variances for 'temperature' ARR and ARRD for both independent and dependent cases. These formulas can also be applied to changes in other factors (e.g., salinity, pH, oxygen) by changing the temperature constant. For example, it may be possible that males and females were measured in slightly different temperatures; say females at $T_3$ and $T_4$ ($T_4 > T_3$) and male at $T_5$ and $T_6$ ($T_6 > T_5$) like below:

@@

Calculating ARR assumes that the units for the mean difference across studies is consistent. Frequently, this will not be the case for meta-analyses that hope to synthesise a wide array of traits that vary in their units (e.g., ul/g, grams, seconds, minutes). Fortunately, $lnRR$ and SMD (Table 1) already controls for unitary differences. As such, we can apply similar logic that we discuss above so that SMD corrects for both temperature and unitary differences. Table 1 provides the formula for SMD, its pooled standard deviation, and sampling variance. Just like with ARR, we can apply our temperature correction to the SMD formula as follows:

>>>>

Here, we show how to obtain sampling variance for a slope (or a 'rate of change') when you have measurements of a physiological trait at two points along an environmental gradient (e.g., temperature, salinity, acidity). We then derive sampling variance for the difference between two slopes, and demonstrate how *SMD* can be corrected to account for differences in both units (e.g. cm or mg) and points on an environmental gradient (e.g. temperature of 24 and 30 \&degC). Finally, we introduce a useful approximation technique known as 'the delta method'. The sampling variances associated with these 'new' effect sizes below will allow comparative physiologists to take advantage of powerful meta-analytic models.

@@

Let's start with a real example that comparative physiologists can easily relate with, the slope of responses to temperature acclimation, or as we call it, the acclimation response ratio (*ARR*). *ARR* can be defined as a slope for acclimated physiological responses at two different temperature points, defined as:

@@

Therefore, when $M_{1}$ and $M_{2}$ are independent, we can obtain the sampling variance for *ARR* as:

@@

where $SD_{1}$ and $SD_{2}$ and $N_{1}$ and $N_{2}$ are standard deviations and sampling sizes at temperature $T_{1}$ and $T_{2}$. Readers may find it difficult to see how we obtained this equation. Let us explain further. The (sampling) standard error for $X_{1}$ is $SE_{1} = SD_{1} / \sqrt{N_{1}}$. Given this, the sampling variance for $X_{1}$ is $SD^2_{1} / N_{1}$, and the sampling variance for $X_{2}$ is $SD^2_{2} / N_{2}$. You can see those elements of the sampling variance of *ARR* in the formula above. The term $(1/(T_{2} - T_{1}))^2$ comes from recognizing $1/(T_{2} - T_{1})$ as a constant.

In ecological and evolutionary studies, it is not uncommon that we measure the same group of organisms at two temperatures. If we do, however, then we need to add the covariance between $M_{1}$ and $M_{2}$ ($\text{Cov}(M_{1},M_{2})$) in the equation. Note that, as above, the covariance equals the correlation multiplied by the square-root of two (sampling) variances. Therefore, the sampling variance of *ARR* can be now written as:

@@

Now let us assume that we want to know the difference between two different *ARR* values: female *ARR*, $\text{ARR}_{f}$ and male *ARR*, $\text{ARR}_{m}$. Such a difference (*ARRD*) can be written as:

@@

Using the properties of variance and the equations from above, the sampling variance for *ARRD* can be derived, when measurements at two temperature points are independent, as:

@@

Here we have derived sampling variances for 'temperature' *ARR* and *ARRD* for both independent and dependent cases. These formulas can also be applied to changes in other factors (e.g., salinity, pH, oxygen) by changing the temperature constant. For example, it may be possible that males and females were measured in slightly different temperatures; say females at $T_3$ and $T_4$ ($T_4 > T_3$) and male at $T_5$ and $T_6$ ($T_6 > T_5$) like below:

@@

Calculating *ARR* assumes that the units for the mean difference across studies is consistent. Frequently, this will not be the case for meta-analyses that hope to synthesise a wide array of traits that vary in their units (e.g., ul/g, grams, seconds, minutes). Fortunately, *$lnRR$* and *SMD* (Table 1) already controls for unitary differences. As such, we can apply similar logic that we discuss above so that *SMD* corrects for both temperature and unitary differences. Table 1 provides the formula for *SMD*, its pooled standard deviation, and sampling variance. Just like with *ARR*, we can apply our temperature correction to the *SMD* formula as follows:
